Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-09-20 analysis/paper/paper.Rmd
Cleaning repository

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  By developing a methodology for handling a larger number of shoreline dates, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both proxies have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicates a largely diverging development of the two proxies through the Mesolithic. The number of shoreline dated sites undergoes some process of decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. Although this will require further substantiation, we suggest that while not devoid of a demographic signal, mobility patterns is the main determinant for the number of shoreline dated sites considered in the study. Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data is more heavily influenced by population dynamics.
keywords: |
  Mesolithic; Radiocarbon dates; Shoreline dates; Fennoscandia 
# highlights: |
#   These are the highlights. 
---

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

<!-- ## Model comparison and continuous piece-wise linear models -->

<!-- The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models. It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This allows for the identification of key demographic events and a conservative description of population dynamics between these. The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021. -->

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], and re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

```{r tab1}
Quality <- c(1, 2, 3, 4, 5, 6)
Definition <- c("Site delineated by use of a GNSS-device, or a securely georeferenced record. Extensive database entry.", "Secure spatial data. Slight disturbance of the site or somewhat lacking database record.", "Secure spatial data. Damaged site, such as outskirts of a quarry, and/or very limited database entry.", "Surveyed by archaeologists. However, the database entry is extremely limited/unclear, the site geometry is only given\\\\as a point or small automatically generated circle, and/or finds are from the topsoil of a field.", "Likely site but uncertain spatial information. Typical example is recurring stray finds in a field or other larger area.", "Single stray find or unverified claims/suggestions of possible site.")
Count <- c(nrow(filter(surveyed, quality == 1)), nrow(filter(surveyed, quality == 2)), nrow(filter(surveyed, quality == 3)), nrow(filter(surveyed, quality == 4)),
           nrow(filter(surveyed, quality == 5)), nrow(filter(surveyed, quality == 6)))

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5 and 97.5 quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline-dated sites, where the probability of drawing any 5-year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *ADMUR*.

<!-- ## Continuous piece-wise linear models -->
<!-- The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021. -->

@@

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

<!-- Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates. -->

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

<!-- For the SSPD all of the standard models could be rejected. The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model. An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge. A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material. Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model. The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.      -->

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs. Lower values indicate a better model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2017; @bjerck2008; @fuglestvedt2012]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs sometime around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.

<!-- @berg-hansen2022 found an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. Their analysis covers a more limited temporal span and geographical area, and is based on the aggregation of point-based estimates of shoreline dates within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. However, consistent with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and, although with a smaller magnitude, around 7100--6900 BCE. -->

@@

Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects. Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

@@

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites is heavily influenced by mobility patterns. Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies indicate that they do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to lose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v.1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

```{r tab1}
Quality <- c(1, 2, 3, 4, 5, 6)
Definition <- c("Site delineated by use of a GNSS-device, or a securely georeferenced record. Extensive database entry.", "Secure spatial data. Slight disturbance of the site or somewhat lacking database record.", "Secure spatial data. Damaged site, such as outskirts of a quarry, and/or very limited database entry.", "Surveyed by archaeologists. Database entry is extremely limited/unclear, the site geometry is only given\\\\as a point or small automatically generated circle, and/or finds are from the topsoil of a field.", "Likely site but uncertain spatial information. Typical example is recurring stray finds in a field or other larger area.", "Single stray find or unverified claims/suggestions of possible site.")
Count <- c(nrow(filter(surveyed, quality == 1)), nrow(filter(surveyed, quality == 2)), nrow(filter(surveyed, quality == 3)), nrow(filter(surveyed, quality == 4)),
           nrow(filter(surveyed, quality == 5)), nrow(filter(surveyed, quality == 6)))

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10,000 times for each null model. The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline-dated sites, where the probability of drawing any 5-year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10,000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *ADMUR*.

@@

For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

All the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occur with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs. Lower values indicate a better model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @åstveit2014; @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2008; @bjerck2017; @fuglestvedt2012]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns starts around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects. Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

@@
